The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-10 rnw/Twophase_Estimators and their Application.Rnw
update Andreas 10.08.2017(1)

In order to apply the \textbf{non-exhaustive two-phase estimation with boundary weight adjustment} to the \code{grisons} data, we have to specify a) the regression model (\code{formula}) as specified in the \code{lm}-function \citep{R}, b) the data frame containing the inventory information (\code{data}), c) the name of the column specifying to which sample ($s_1$, $s_2$) an observation belongs (\code{phase.col}) as well as the indicator for the $s_2$, i.e. terrestrial sample (\code{terrgrid.id}), and d) the columname of the column containing the weights $w(x)$ of the boundary adjustments. The \textbf{non-exhaustive estimator with boundary weight adjustment} can thus be applied as follows:
\begin{small}
<<>>=
reg2p_nex <- twophase(formula=tvol ~ mean + stddev + max + q75,
                   data=grisons,
                   phase_id=list(phase.col = "phase_id_2p", terrgrid.id = 2),
                   boundary_weights = "boundary_weights")
@
\end{small}
The \code{twophase}-function creates an \code{S3} object of class \code{"global" "twophase"}. A \textbf{summary of the estimation results} can be obtained by applying the \code{summary}-function to that object. It first provides a summary of the function call, the applied regression formula and the estimation method that was applied. The estimation results comprise the point estimate (\code{estimate}) as well as the design-based and external variance (\code{g_variance} and \code{ext_variance}). Further given are the sample sizes of the $s_1$ and $s_2$ sample (\code{n1}, \code{n2}) as well as the R$^2$ of the regression model.
\begin{small}
<<>>=
summary(reg2p_nex)
@
\end{small}
If the exact means of the explanatory variables $\bar{\pmb{Z}}$ are available, the \textbf{exhaustive estimator} can be applied by additionally passing a vector containg the exact means to the optional argument \code{exhaustive}. The vector must be input in the same order that the \code{lm}-function processes a \code{formula} object including the intercept term whose exact mean will always be 1. We assume that in this case, the boundary adjustment has been applied when calculating the exact mean of $\pmb{Z}$.

@@

We first introduce the \textbf{small area estimator} (\textit{small}), which uses exhaustively computed explanatory variables, and its non-exhaustive version, the \textbf{\textit{pseudo}} \textbf{small area estimator} (\textit{psmall}). Their point estimates are given in equation \ref{eq:pointest_2p_reg_small} and \ref{eq:pointest_2p_reg_psmall}. As obvious, the globally derived regression coefficients are first applied to the exhaustively or non-exhaustively calculated means of the explanatory variables available at the sample location $s_{1,G}$ in the small area. A potential bias of the regression model predictions in the small area $G$ is then corrected by adding the mean of the residuals in $G$ to the first term. We call this the \textit{bias} or \textit{residual correction} term.

@@

The package again provides the \textit{design-based} (equation \ref{eq:var_2p_reg_small}, \ref{eq:var_2p_reg_psmall}) and \textit{external} variance (equation \ref{eq:varext_2p_reg_small}, \ref{eq:varext_2p_reg_psmall}) for the \textbf{small area estimator} and the \textbf{pseudo small area estimator}. Again note that all components are restricted those available at the sample locations $s_{1,G}$ and $s_{2,G}$ in the small area, with exception of the regression coeffient components $\hat{\pmb{\beta}}_{s_2}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$.

@@

@@

The \textbf{synthetic} and \textbf{pseudo synthetic estimator} is applied if no terrestrial samples are available within the small area $G$ (i.e. $n_{2,G}=0$). In this case, the point estimate (equation \ref{eq:pointest_2p_reg_synth} and \ref{eq:pointest_2p_reg_psynth}) is based \textit{only} on the predictions generated by applying the globally derived regression model to the auxiliary vector $\bar{\pmb{Z}}_G$ or $\hat{\bar{\pmb{Z}}}_G$. However, the bias correction applied in the small and pseudo small area estimator (equation \ref{eq:pointest_2p_reg_small} and \ref{eq:pointest_2p_reg_psmall}) is no longer possible. Thus, the synthetic / pseudo synthetic estimator is unbiased \textbf{only} if the \textit{zero mean property} of the residuals $\int_F R(x)dx=0$ also holds for the residuals in the small area $G$, i.e. $\int_G R(x)dx=0$. If this assumption is not fullfilled, the synthetic point estimate $\hat{Y}_{G,synth,2p}$ and the pseudo synthetic point estimate $\hat{Y}_{G,psynth,2p}$ will have a design-based asymptotic bias equal to $-\frac{1}{\lambda(G)}\int_G R(x)$. Also note that residual variation can no longer be considered in the design-based variance (equation \ref{eq:var_2p_reg_synth} and \ref{eq:var_2p_reg_psynth}). Therefore, the synthetic estimators will usually have a smaller variance than estimators incorporating the regression model uncertainties, but at the cost of a potential bias. Due to the absence of available residuals in $G$, there is also no external variance form for synthetic estimations.

@@

The \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator}

% - explain extended estimator
% - mention boundary weight adjustment likewise global estimation

@@

In order to apply the \textbf{pseudo small area estimator} (\textit{psmall}), we set \code{unbiased=TRUE} as well as the optional argument \code{psmall=TRUE}:
\begin{small}
<<>>=
psmall_2p <- twophase(formula = tvol ~ mean + stddev + max + q75, data = grisons,
                      phase_id = list(phase.col = "phase_id_2p", terrgrid.id = 2),
                      boundary_weights = "boundary_weights",
                      small_area = list(sa.col = "smallarea", areas = c("A", "B"),
                                        unbiased = TRUE),
                      psmall = TRUE)
summary(psmall_2p)
@
\end{small}

The small area functions all return an \code{S3} object of class \code{"smallarea" "twophase"}. In addition to global estimation, the \code{summary} output now also contains the sample sizes of each small area (\code{n1G} and \code{n2G}). Note that this information is also particularly provided in the \code{samplesizes} object returned by the function call.
\begin{small}
<<>>=
psmall_2p$samplesizes
@
\end{small}

@@

Comparing this result with the non-exhaustive counterpart \textit{extpsynth}, we see that the estimation uncertainty (defined by the variance) has again be significantly increased by substituting the estimated by the exact auxiliary means. In the function output, exhautive phases are indicated by a sample size of infinity \code{Inf}. Also note that the function extracts the required exact means for small area \code{"A"} and \code{"B"} from the complete set \code{true.means.Z.G}.

>>>>

In order to apply the \textbf{non-exhaustive two-phase estimation with boundary weight adjustment} to the \code{grisons} data, we have to specify a) the regression model (\code{formula}) as specified in the \code{lm}-function \citep{R}, b) the data frame containing the inventory information (\code{data}), c) the name of the column specifying to which sample ($s_1$, $s_2$) an observation belongs (\code{phase.col}) as well as the indicator for the $s_2$, i.e. terrestrial sample (\code{terrgrid.id}), and d) the columname of the column containing the weights $w(x)$ of the boundary adjustments. The \textbf{non-exhaustive estimator with boundary weight adjustment} can thus be applied as follows:
\begin{small}
<<>>=
reg2p_nex <- twophase(formula=tvol ~ mean + stddev + max + q75,
                   data=grisons,
                   phase_id=list(phase.col = "phase_id_2p", terrgrid.id = 2),
                   boundary_weights = "boundary_weights")
@
\end{small}
The \code{twophase}-function creates an \code{S3} object of class \code{"twophase"} and subclass \code{"global"}. A \textbf{summary of the estimation results} can be obtained by applying the \code{summary}-function to that object. It first provides a summary of the function call, the applied regression formula and the estimation method that was applied. The estimation results comprise the point estimate (\code{estimate}) as well as the design-based and external variance (\code{g_variance} and \code{ext_variance}). Further given are the sample sizes of the $s_1$ and $s_2$ sample (\code{n1}, \code{n2}) as well as the R$^2$ of the regression model (\code{r.squared}).
\begin{small}
<<>>=
summary(reg2p_nex)
@
\end{small}
If the exact means of the explanatory variables $\bar{\pmb{Z}}$ are available, the \textbf{exhaustive estimator} can be applied by additionally passing a vector containg the exact means to the optional argument \code{exhaustive}. The vector must be input in the same order that the \code{lm}-function processes a \code{formula} object including the intercept term whose exact mean will always be 1. We assume that in this case, the boundary adjustment has been applied when calculating the exact mean of $\pmb{Z}$.

@@

We first introduce the \textbf{small area estimator} (\textit{small}), which uses exhaustively computed explanatory variables, and its non-exhaustive version, the \textbf{\textit{pseudo}} \textbf{small area estimator} (\textit{psmall}). Their point estimates are given in equation \ref{eq:pointest_2p_reg_small} and \ref{eq:pointest_2p_reg_psmall}. As obvious, the globally derived regression coefficients are first applied to the exhaustively or non-exhaustively calculated means of the explanatory variables ($\bar{\pmb{Z}}_G$, $\hat{\bar{\pmb{Z}}}_G$) available at the sample location $s_{1,G}$ in the small area. A potential bias of the regression model predictions in the small area $G$ is then corrected by adding the mean of the residuals in $G$ to the first term. We call this the \textit{bias} or \textit{residual correction} term.

@@

The package again provides the \textit{design-based} (equation \ref{eq:var_2p_reg_small}, \ref{eq:var_2p_reg_psmall}) and \textit{external} variance (equation \ref{eq:varext_2p_reg_small}, \ref{eq:varext_2p_reg_psmall}) for the \textit{small} and the \textit{psmall} estimator. Again note that all components are restricted to those available at the sample locations $s_{1,G}$ and $s_{2,G}$ in the small area, with exception of the regression coeffient components $\hat{\pmb{\beta}}_{s_2}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$.

@@

If boundary adjustment is applied, the simple mean of the explanatory variable over the small area $\hat{\bar{\pmb{Z}}}_G=\frac{1}{n_{1,G}}\sum_{x \in s_{1,G}}\pmb{Z}(x)$ is replaced by its weighted version $\hat{\bar{\pmb{Z}}}_G=\frac{\sum_{x\in{s_{1,G}}}w(x)\bmp{Z}(x)}{\sum_{x\in{s_{1,G}}}w(x)}$, and likewise for exhaustively used auxiliary information.

@@

The \textbf{synthetic} and \textbf{pseudo synthetic estimator} is applied if no terrestrial samples are available within the small area $G$ (i.e. $n_{2,G}=0$). In this case, the point estimate (equation \ref{eq:pointest_2p_reg_synth} and \ref{eq:pointest_2p_reg_psynth}) is based \textit{only} on the predictions generated by applying the globally derived regression model to the auxiliary vector $\bar{\pmb{Z}}_G$ or $\hat{\bar{\pmb{Z}}}_G$. However, the bias correction applied in the small and pseudo small area estimator (equation \ref{eq:pointest_2p_reg_small} and \ref{eq:pointest_2p_reg_psmall}) is no longer possible. Thus, the synthetic / pseudo synthetic estimator is unbiased \textbf{only} if the \textit{zero mean property} of the residuals $\int_F R(x)dx=0$ also holds for the residuals in the small area $G$, i.e. $\int_G R(x)dx=0$. If this assumption is not fullfilled, the synthetic point estimate $\hat{Y}_{G,synth,2p}$ and the pseudo synthetic point estimate $\hat{Y}_{G,psynth,2p}$ will have a design-based asymptotic bias equal to $-\frac{1}{\lambda(G)}\int_G R(x)$. Also note that residual variation can no longer be considered in the design-based variance (equation \ref{eq:var_2p_reg_synth} and \ref{eq:var_2p_reg_psynth}). Therefore, the synthetic estimators will usually have a smaller variance than estimators incorporating the regression model uncertainties, but at the cost of a potential bias. Due to the absence of available residuals in $G$, there is also no external variance form for the synthetic and pseudo synthetic estimator.

@@

In the \textit{small} and \textit{psmall} estimator, the crucial zero mean residual property for the small area $G$ is forced by the bias adjustment terms (equation \ref{eq:pointest_2p_reg_small}, \ref{eq:pointest_2p_reg_psmall}). However, the equations would considerably simplify if we could mathematically insure that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is idea is realized in the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) by extending the vector of auxiliary information $\pmb{Z}}(x)$ by a binary categorical indicator variable which has the value \code{1} if the observation lies inside the small area $G$ and otherwise set to \code{0}. The \textit{extended} auxiliary vector thus reads as $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))\in{\mathbb{R}}^{(p+1)}$ and corresponds to the classical ANOVA case. Note that this adjusts the intercept of the regression model for all data points of small area $G$ and thus insures that the mean of the residuals $\frac{1}{n_{2,G}}\sum_{x\in{s_{2,G}}}\mathbb{R}(x)$ in $G$ are always zero by construction.\par

Under the extended auxiliary vector, the regression coefficient .. is calculated as

\begin{equation}\label{ext_normequ_simple}
  \hat{\pmb{\theta}}_{s_2}&=& \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1} \Big(\frac{1}{n_2}\sum_{x\in{s_2}}Y(x)\pmb{\mathbb{Z}}(x)\Big)
\end{equation}

with its variance-covariance matrix

\begin{equation}\label{eq:ext_estvarmatrix}
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}:=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{\mathbb{Z}}(x)\pmb{Z}(x)^t\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
\end{equation}

The point estimates are defined as

\begin{subequations}\label{eq:pest_2p_extsynth_extpsynth}
\begin{align}
\hat{Y}_{G,extsynth}= \bar{\pmb{\mathbb{Z}}}^t_{G}\hat{\pmb{\theta}}_{s_2} \label{eq:pointest_2p_extsynth} \\
\hat{Y}_{G,extpsynth}=\hat{\bar{\pmb{\mathbb{Z}}}}_{G}^t\hat{\pmb{\theta}}_{s_2} \label{eq:pointest_2p_extsynth}
\end{align}
\end{subequations}

The estimators are called \textit{synthetic} since no residual term is present such as in the small and psmall estimator (equation ...) and thus look similar to the synthetic estimators (equation ...). Note that it can however be shown by a decomposition of $\hat{\theta}_{s2}$ that the residual correction is indeed included in the regression coefficient $\hat{\theta}_{s2}$ \citep{mandallaz2016}.

@@

In order to apply the \textbf{pseudo small area estimator} (\textit{psmall}) with boundary adjustment, we set \code{unbiased=TRUE} as well as the optional argument \code{psmall=TRUE}:
\begin{small}
<<>>=
psmall_2p <- twophase(formula = tvol ~ mean + stddev + max + q75, data = grisons,
                      phase_id = list(phase.col = "phase_id_2p", terrgrid.id = 2),
                      boundary_weights = "boundary_weights",
                      small_area = list(sa.col = "smallarea", areas = c("A", "B"),
                                        unbiased = TRUE),
                      psmall = TRUE)
summary(psmall_2p)
@
\end{small}

The small area functions all return an \code{S3} object of class \code{"twophase"} and subclass \code{"smallarea"}. In addition to global estimation, the \code{summary} output and the \code{estimation} object comprise the estimates and the sample sizes (\code{n1G} and \code{n2G}) for all calculated small areas (column \code{area}). Note that the sample size information is also particularly provided in the \code{samplesizes} object returned by the function call.
\begin{small}
<<>>=
psmall_2p$samplesizes
@
\end{small}

@@

Comparing this result with the non-exhaustive counterpart \textit{extpsynth}, we see that the estimation uncertainty (defined by the variance) has again be significantly increased by substituting the estimated by the exact auxiliary means. In the function output, exhautive phases are indicated by a sample size of infinity (\code{Inf}). Also note that the function extracts the required exact means for small area \code{"A"} and \code{"B"} from the complete set of exact means defined in \code{true.means.Z.G}.
